sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-11 doc/appendix-snow-sdmTMB.Rmd

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

```{r load-data, cache=FALSE} 
snow <- readRDS(here::here("owls/SNOW_data.rds"))
glimpse(snow)
```

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)`.

@@

```{r plot-mesh}

```

@@

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

@@

Print the model to inspect the fit:
```{r check, include = FALSE, cache=FALSE, eval=TRUE}
m # loading this twice just prevents some excess messages from cluttering the output when printing below
```

@@

```{r stan-resids, eval=FALSE}
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "mcmc_residuals.rds")
```

```{r stan-dat, echo=FALSE, cache=FALSE}
r_mcmc <- readRDS(here::here("owls/mcmc_residuals.rds"))
```

@@

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transformed them to the correct projection.

```{r shapes}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

>>>>

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's Climatic Research Unit ([ https://crudata.uea.ac.uk/cru/data/nao/](https://crudata.uea.ac.uk/cru/data/nao/)). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

```{r load-data, cache=FALSE} 
snow <- readRDS(here::here("data/SNOW_data.rds"))
glimpse(snow)
```

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)` [@pebesma2018].

@@

@@

```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
```

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
m0 <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
m1 <- readRDS(here::here("data/snow_w_main_effect_1_150km.rds"))
m2 <- readRDS(here::here("data/snow_w_main_effect_150km.rds"))
```

@@

```{r save-reml-model, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
saveRDS(m, here::here("data/snow_w_main_effect_150km_reml.rds"))
```

Print the model to inspect the fit:
```{r check, include = FALSE, cache=FALSE, eval=TRUE}
m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
m # loading this twice just prevents some excess messages from cluttering the output when printing below
```

@@

```{r stan-resids, eval=FALSE}
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "data/mcmc_residuals.rds")
```

```{r stan-dat, echo=FALSE, cache=FALSE}
r_mcmc <- readRDS(here::here("data/mcmc_residuals.rds"))
```

@@

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http//www.naturalearthdata.com/download/10m/physical/ne_10m_lakes.zip) and transformed them to the correct projection.

```{r shapes}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("data/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

### 2021-12-11 doc/paper.Rmd

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

>>>>

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
